\pdftitle

### 2017-05-17 maxNt.tex
intro

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T21:23:29+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varIota'}
\newcommand*{\yHb}{\varIota''}
\newcommand*{\yHc}{\varKappa}
%Lagr multipliers
\newcommand*{\yL}{\varLambda}
\newcommand*{\yl}{\lambda}
\newcommand*{\yk}{\kappa}
\newcommand*{\yK}{\varKappa}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

Recent electrophysiological techniques in fact allow experimenters to
record samples of even a couple hundreds neurons from specific brain areas.
We don't elaborate on the various important purposes of such recordings
here, but stress that these sample sizes can be considered \enquote{large},
because their statistical analysis requires considerable computational
power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. Maximum-entropy methods make this kind of compression:
they associate a unique probability distribution with few experimental
quantities. This distribution is usually said to be \enquote{maximally
  noncommittal} \cite{jaynes1963}.

We intend the quoted expression only as
an umbrella term, also because it means very little without further
clarifications.

@@

Suppose we want to constrain the sample expectations of a vector function
$f$ to the vector values $c$, that is, $\expeb{f(\yxs)}=c$. Application of
\me\ at the population level, denoted by $\yHa$, gives
\begin{equation}
  \label{eq:app_maxent_pop}
  \p(\yXf = \yRf \| \yHa)  = \yK\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
and then by marginalization with \eqn~\eqref{eq:marginal}
\begin{equation}
  \label{eq:app_maxent_pop_marg}
  \p(\yxs = \yrs \| \yHa)  = \yK\, 
  \sum_{N\yRf=0}^N \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
where $\yK$ is a normalization constant and $\yL\T$ are Lagrange
multipliers such that
\begin{equation}
  \label{eq:app_maxent_pop_constraints}
c = \yK\sum_{n\yrs=0}^n \sum_{N\yRf=0}^N f(\yrs) \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yRf)\Bigr].
\end{equation}
Application of \me\ at the sample level, denoted by $\yHb$, gives
\begin{equation}
  \label{eq:app_maxent_sample}
  \p(\yxs=\yrs \|\yHb)
  =\yk\,
  \binom{n}{n\yrs}
  \exp[\yl\T f(\yrs)]
\end{equation}
where $\yk$ is a normalization constant and $\yl\T$ are Lagrange
multipliers such that
\begin{equation}
  \label{eq:app_maxent_sample_constraints}
c = \yk\sum_{n\yrs=0}^n  f(\yrs) \binom{n}{n\yrs}
  \exp[\yl\T f(\yrs)].
\end{equation}

The probabilities for the sample average obtained from application at the
population level~\eqref{eq:app_maxent_pop_marg} and at the sample level
\eqref{eq:app_maxent_sample} will be approximately equal, by our previous
observation~\eqref{eq:roughly_equal_nN} and also by the fact that they must
satisfy the same expectations for $f$. But the latter may miss important
features present in the former, like additional modes or particular tail
behaviour.

@@

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T22:40:04+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\yxx}{x}%subpop state
\newcommand*{\yx}{\bm{\yxx}}%subpop state
\newcommand*{\yxs}{\sav{\yx}}%subpop av state
\newcommand*{\yX}{\bm{X}}%pop state
\newcommand*{\yy}{\bm{y}}%pop state
\newcommand*{\yXf}{\av{\yX}}%pop av state
\newcommand*{\yxxs}{\sav{\yx\yx}}%subpop av state
\newcommand*{\yXXf}{\av{\yX\yX}}%subpop av state
\newcommand*{\yr}{\bm{r}}%subpop value
\newcommand*{\ys}{\bm{s}}%subpop value
\newcommand*{\yrs}{\sav{\yr}}%subpop av value
\newcommand*{\yR}{\bm{R}}%pop value
\newcommand*{\yRf}{\av{\yR}}%pop av value
%conditional assumptions
\newcommand*{\yH}{\varIota}
\newcommand*{\yD}{\varDelta}
\newcommand*{\yHa}{\varIota'}
\newcommand*{\yHb}{\varIota''}
\newcommand*{\yHc}{\varKappa}
%Lagr multipliers
\newcommand*{\yf}{\bm{f}}
\newcommand*{\yc}{\bm{c}}
\newcommand*{\yL}{\bm{\varLambda}}
\newcommand*{\yl}{\bm{\lambda}}
\newcommand*{\yk}{\kappa}
\newcommand*{\yK}{\varKappa}
%entropies
\newcommand*{\ysh}{H}
\newcommand*{\ybu}{H_{\text{B}}}

@@

Recent electrophysiological techniques \cite{***} in fact allow
experimenters to record samples of even a couple hundreds neurons from
specific brain areas. These samples are usually picked out according to an
unknown process, but from their observation we expect to learn something
about all other neurons in the same brain area. That is, we assumed
that they are a \enquote{representative random sample}.

We don't elaborate on the various important purposes of such recordings
here \cite{?***}, but stress that these sample sizes can be considered
\enquote{large}, because their statistical analysis requires considerable
computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. The standard \me\ method
\cite{jaynes1954_r1974,sivia1996_r2006,meadetal1984} achieve this kind of
compression: they associate a unique probability distribution with few
experimental quantities \cite{***}. But this is only one of their uses.
They are also used for various information-theoretic purposes or to
generate reference probability distributions \cite{***}. In all these uses
the \me\ distribution is chosen as the \enquote{maximally noncommittal} one
\cite{jaynes1963}. This adjective means little without further technical
characterizations. Different works give different characterizations, and in
this paper we will use the quoted expression as an umbrella term for all of
them. Our results will not depend on the specific characterization of
\enquote{noncommittal}.

In this note we want to show that we have to face a dilemma if we want to
apply the \me\ method to a representative random sample to find a maximally
noncommittal distribution.

The dilemma is this. We can apply the \me\ method to the sample, using a
specified set of experimental constraints, and generate a probability
distribution for its state. But our sample is representative of a larger
population. We can apply the \me\ method to the larger population, using
the same constraints, and generate a probability distribution for its
larger state, and then find the distribution for the sample by
marginalization. Either application seems to have some commendable
features. However, \emph{the distributions obtained by these two
  applications differ}. It goes without saying that if one is
\enquote{maximally noncommittal}, the other must be somehow
\enquote{committal}. Which choice is most meaningful?

In the rest of the paper we mathematically formulate this dilemma. To this
purpose we also present some probability relations relevant to
\enquote{random sampling}. These relations are well-known in survey
sampling and in the pedagogic problem of drawing from an urn without
replacement, yet they are somewhat hard to find explicitly written in the
neuroscientific literature, so they may be of interest on their own.

@@

Suppose we want to constrain the sample expectations of a vector function
$\yf = (f_1,\dotsc,f_m)$ to the vector values $\yc=(c_1,\dotsc,c_m)$, that
is, $\expeb{\yf(\yxs)}=\yc$. Application of \me\ at the population level,
denoted by $\yHa$, gives
\begin{equation}
  \label{eq:app_maxent_pop}
  \p(\yXf = \yRf \| \yHa)  = \yK\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
and then by marginalization with \eqn~\eqref{eq:marginal}
\begin{equation}
  \label{eq:app_maxent_pop_marg}
  \p(\yxs = \yrs \| \yHa)  = \yK\, 
  \sum_{N\yRf=0}^N \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
where $\yK$ is a normalization constant and
$\yL\T=(\varLambda_1,\dotsc,\varLambda_m)\T$ are Lagrange multipliers such
that
\begin{equation}
  \label{eq:app_maxent_pop_constraints}
\yc = \yK\sum_{n\yrs=0}^n \sum_{N\yRf=0}^N \yf(\yrs) \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr].
\end{equation}
Application of \me\ at the sample level, denoted by $\yHb$, gives
\begin{equation}
  \label{eq:app_maxent_sample}
  \p(\yxs=\yrs \|\yHb)
  =\yk\,
  \binom{n}{n\yrs}
  \exp[\yl\T \yf(\yrs)]
\end{equation}
where $\yk$ is a normalization constant and $\yl\T$ are Lagrange
multipliers such that
\begin{equation}
  \label{eq:app_maxent_sample_constraints}
\yc = \yk\sum_{n\yrs=0}^n  \yf(\yrs) \binom{n}{n\yrs}
  \exp[\yl\T \yf(\yrs)].
\end{equation}

The probabilities for the sample average obtained from application at the
population level~\eqref{eq:app_maxent_pop_marg} and at the sample level
\eqref{eq:app_maxent_sample} will be approximately equal, by our previous
observation~\eqref{eq:roughly_equal_nN} and also by the fact that they must
satisfy the same expectations for $\yf$. But the latter may miss important
features present in the former, like additional modes or particular tail
behaviour.

@@

**********************************************************
